Modal Abstractions for Virtualizing Memory Addresses

### 2024-07-04 examples.tex
+

%\begin{comment}
%\todo[inline]{Identity mappings are difficult, and our current approach won't quite work. Consider trying to have a virtual pointsto for an actual page table entry (i.e., that one could use to update a page table mapping), while also having a virtual pointsto for an address that entry mapped. With the current (let's call it v1) solution, we can't actually have both of those simultaneously!  That's because the PTE pointsto will assert full ownership of the physical memory cell holding the PTE as its data value, while the virtual pointsto for the data mapped by that entry will \emph{also} assert (fractional) ownership of all entries a page table walk would traverse.
%}
%\todo[inline,color=violet]{This doesn't seem to cause issues with the mapping/unmapping examples, only with changing intermediate page table pointers. The mapping example requires a virtual pointsto for the blank PTE, and once filled in that ownership can be immediately split to create the 512 new virtual pointsto assertions for the newly mapped page. Conversely, for unmapping we'd assume ownership of all the relevant virtual pointsto assertions for the page we're unmapping, at which point we can (with a bit of work) show that they all correspond to the same L1 PTE, and extract the 512 fractional shares of that entry from the pointsto assertions.  But changing intermediate page tables, as one would do for coallescing or splitting a superpage while preserving the virtual-to-physical mappings, couldn't be done without some really complicated separating implication tricks.}
%\todo[inline,color=green]{One possible approach to resolving this, which we came up with in our Tuesday meeting, is to recognize that the current (v1) virtual points-to is too strong, because it really doesn't care about \emph{owning} those fractional resources, it only cares that \emph{something} ensures the correct page table walk exists. Iris has a ghost map resource where authoritative ownership of an individual key-value pair can be handled as a resource.  (Colin was using this in the filesystem cache.)
%We can use that mechanism to separate the virtual-to-physical translation from the physical memory involved (Kolanski and Klein may have done something similar for different reasons): (fractional) virtual points-to assertions can be defined in terms of (fractional) ownership of these authoritative ghost map entry assertions, plus sharing an invariant that the current installed page table respects all entries of the mapping. Unmapping collects the authoritative map kvpairs from collecting the assertions, and then can remove them from the ghost map and update the page tables. Critically, physical ownership of the page tables then lives in the invariant on the current page table, so some virtual pointsto assertions can refer to memory in those page tables.
%This still works with the modality, since that invariant is also semantically a predicate on a page table root.
%Let's call this v2.
%}
%\end{comment}
\subsection{Mapping a New Page}
One of the key tasks of a page fault handler in a general-purpose OS kernel is
to map new pages into an address space by writing into an existing page table (\textsf{vaspace\_mappage(pte\_t *pml4, void *va,uintptr\_t pa, uint64\_t perm )} in Figure \ref{fig:mapping_code}).
To do so, with a given allocated a fresh page (\textsf{pa}), then calculate the appropriate
known-valid page table walks (\textsf{pte\_walkpgdir} Line X in Figure \ref{fig:mapping_code})  and update the appropriate L1 page table entry (permissions and frame address Lines in Figure \ref{fig:mapping_code}).
%\lstset{
%  columns=fullflexible,
%  numbers=left,
%  basicstyle=\ttfamily,
%  keywordstyle=\color{blue}\bfseries,
%  morekeywords={mov,add,call},
%  emph={rsp,rdx,rax,rbx,rbp,rsi,rdi,rcx,r8,r9,r10,r11,r12,r13,r14,r15},
%  emphstyle=\color{green},
%  emph={[2]cr3},
%  emphstyle={[2]\color{violet}},
%  morecomment=[l]{;;},
%  mathescape
%}
\newcommand{\sumwalkabsent}{
  \ownGhost\gammaPred{\authfrag{\singletonMap{\texttt{entry+KERNBASE}}{(\textsf{qfrac}, \textsf{entry})}}}

\newcommand{\fpaddr}{\texttt{fpaddr}}
\newcommand{\specline}[1]{{\color{blue}\left\{#1\right\}}}
\newcommand{\sumapacesfull}[2]{
  \ownGhost\gammaPreds{\authfull{\singletonMap{#1}{#2}}}
}
\todo{ismail change the following spec base on the changes you have in the code}
\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape]
;;complstatus_t vaspace_mappage(pte_t *pml4, void *va,uintptr_t pa, uint64_t perm ) {
...
;;setting up the stack      
;;char *a; pte_t *pte;
;;a = (char *)PGROUNDUP((uintptr_t)va);
 $\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,m)  \ast  \texttt{r14}\mapsto_{\textsf{r}} \_ \ast \texttt{rdi}\mapsto_{r} \vaddr \ast \texttt{rax}\mapsto_{\textsf{r}} \_ \ast \ulcorner \texttt{aligned va} \land  \theta \; !!\; \vaddr = \texttt{None}\urcorner}_{\rtv}$
;;char *a; pte_t *pte;
;;a = (char *)PGROUNDUP((uintptr_t)va);
mov    -0x10[rbp],rcx
add    0x1000,rcx
sub    0x1,rcx
and    0xfffffffffffff000,%rcx
mov    rcx,-0x28[rbp]

;;pte = walkpgdir_succ(pml4, a);
mov    -0x8[rbp],rdi
mov    -0x28[rbp],rsi
mov    0x0,rax
callq  291 <vaspace_mappage+0x41>	28d: R_X86_64_PLT32	walkpgdir_succ-0x4
mov    0x0,r8
movslq eax,rcx
mov    rcx,-0x30[rbp]
;;pte->writable     = PTE_WRITEABLE_SET(perm);
...
pte->user_acc     = PTE_USER_ACC_SET(perm);
...
;;pte->writethrough = PTE_WRITETHRU_SET(perm);
...
;;pte->nocache      = PTE_NOCACHE_SET(perm);
pte->pfn          = PTE_ADDR_TO_PFN(pa);
mov    -0x18[rbp],rcx
shr    0xc,rcx
mov    -0x30[rbp],rdx
mov    [rdx],rsi
movabs 0xffffffffff,rdi
and    rdi,rcx
shl    0xc,rcx
movabs 0xfff0000000000fff,rdi
and    rdi,rsi
or     rcx,rsi
mov    rsi,[rdx]
;;pte->nx           = PTE_NX_SET(perm);
;;pte->present      = 1;
;;clean up the stack and return
...
;;}
\end{lstlisting}
  \caption{Specification and proof of code for mapping a new page.}
\label{fig:mapping_code}
\end{figure}

In Figure \ref{fig:mapping_code}, we see an address ($\vaddr$) currently not
mapped to a page ($\theta \; !!\; \vaddr = \texttt{None}$). Mapping a fresh
phyiscal page to back the desired virtual page first requires ensuring
the existence of a memory location for an appropriate L1 table entry.
The code uses a helper function \lstinline|pte\_walkpgdir| (discussed again in Section \ref{sec:traversing}).
\textsf{pte\_walkpgdir}'s postcondition contains virtual \emph{PTE} pointsto assertions ($\mapsto_{\textsf{vpte}}$) \todo{colin, these are no more plain physical pointstos}  both for ensuring partial page table walk reaching the
L1 entry (l1e) with the justification of higher levels of the page table exist, and for allowing access to the memory of the L1 entry via virtual address.
A PTE points-to is defined just like the normal virtual points-to of Figure \ref{fig:virtualpointstosharing}, except the physical address (here, \textsf{pa}) is explicit in the assertion 
rather than existentially quantified:
 \[
\begin{array}{l}
    \vaddr\mapsto_{\textsf{vpte,q}} \; \paddr \; \vpage : \mathsf{vProp}~\Sigma \stackrel{\triangle}{=} 
    \exists \delta\ldotp
	(\lambda \mathit{cr3val}\ldotp
	\sumapaces{\mathit{cr3val}}\delta) \ast 
  \sumwalkabs\vaddr\qfrac\paddr \ast \paddr \mapsto_{\mathsf{p}} \vpage
\end{array}
\]
This supports rules for accessing memory
at that virtual address, but exposing the physical location being modified
makes this convenient to use for page table modifications, since we must ensure
the modified data is at the specific physical location that will be used as the L1 entry.
This is also guaranteed ($ \ulcorner
\texttt{addr\_L1}(\vaddr,\entryo) = \paddr\urcorner$).
% After obtaining a virtual address \textsf{pte\_addr} in \textsf{rax} backed 
% by the physical memory for the L1 entry that will be used to translate the virtual addresses
% we are mapping, we save it to \textsf{r14} to be updated later in Line 9.

The crucial step is actually updating the L1 entry,
via the virtual address
(\textsf{pte\_addr}) known to translate to the appropriate physical address, in our example the L1
table entry address ($\textsf{addr\_L1}(\textsf{va, l1e})$), to hold the freshly
allocated physical page address (\textsf{fpaddr}) in Line X.
After this write, we have (Specification Line Y)
\[\texttt{pte\_addr} \mapsto_{\texttt{vpte}}  \; \paddr\; \textsf{fpaddr}  \qquad\qquad (\textsf{by the PTE variant of Rule}~\TirNameStyle{WriteToVirtMemFromReg})\]

\subsection{Unmapping a Page}
The reverse operation, unmapping a designated page that is currently mapped,
would essentially be the reverse of
the reasoning around line 22 above: given the virtual points-to assertions for all 512
machine words of memory that the L1 entry would map,
and information about the physical location, 
full permission on the L1 entry could be obtained, allowing the construction of a
full virtual PTE pointer for it, setting to 0, and reclaiming the now-unmapped physical memory.

@@

% \textsc{seL4} currently still trusts address translations; it models page tables as a data structure in regular memory, thus not capturing the possibility that even
% temporarily destroying the mappings and restoring them can actually crash the OS. \textsc{CertiKOS} papers share little in the way of precise details about
% their virtual memory management, but because their core technology is based on a fork of \textsc{CompCert}, whose model of memory is
% a set of unordered block allocations, we can infer their proofs must also trust these translations.

>>>>

%\begin{comment}
%\todo[inline]{Identity mappings are difficult, and our current approach won't quite work. Consider trying to have a virtual pointsto for an actual page table entry (i.e., that one could use to update a page table mapping), while also having a virtual pointsto for an address that entry mapped. With the current (let's call it v1) solution, we can't actually have both of those simultaneously!  That's because the PTE pointsto will assert full ownership of the physical memory cell holding the PTE as its data value, while the virtual pointsto for the data mapped by that entry will \emph{also} assert (fractional) ownership of all entries a page table walk would traverse.
%}
%\todo[inline,color=violet]{This doesn't seem to cause issues with the mapping/unmapping examples, only with changing intermediate page table pointers. The mapping example requires a virtual pointsto for the blank PTE, and once filled in that ownership can be immediately split to create the 512 new virtual pointsto assertions for the newly mapped page. Conversely, for unmapping we'd assume ownership of all the relevant virtual pointsto assertions for the page we're unmapping, at which point we can (with a bit of work) show that they all correspond to the same L1 PTE, and extract the 512 fractional shares of that entry from the pointsto assertions.  But changing intermediate page tables, as one would do for coallescing or splitting a superpage while preserving the virtual-to-physical mappings, couldn't be done without some really complicated separating implication tricks.}
%\todo[inline,color=green]{One possible approach to resolving this, which we came up with in our Tuesday meeting, is to recognize that the current (v1) virtual points-to is too strong, because it really doesn't care about \emph{owning} those fractional resources, it only cares that \emph{something} ensures the correct page table walk exists. Iris has a ghost map resource where authoritative ownership of an individual key-value pair can be handled as a resource.  (Colin was using this in the filesystem cache.)
%We can use that mechanism to separate the virtual-to-physical translation from the physical memory involved (Kolanski and Klein may have done something similar for different reasons): (fractional) virtual points-to assertions can be defined in terms of (fractional) ownership of these authoritative ghost map entry assertions, plus sharing an invariant that the current installed page table respects all entries of the mapping. Unmapping collects the authoritative map kvpairs from collecting the assertions, and then can remove them from the ghost map and update the page tables. Critically, physical ownership of the page tables then lives in the invariant on the current page table, so some virtual pointsto assertions can refer to memory in those page tables.
%This still works with the modality, since that invariant is also semantically a predicate on a page table root.
%Let's call this v2.
%}
%\end{comment}
\subsection{Unmapping a Page}
The reverse operation, unmapping a designated page that is currently mapped,
would essentially be the reverse of
the reasoning around line 22 above: given the virtual points-to assertions for all 512
machine words of memory that the L1 entry would map,
and information about the physical location, 
full permission on the L1 entry could be obtained, allowing the construction of a
full virtual PTE pointer for it, setting to 0, and reclaiming the now-unmapped physical memory.

@@

% \textsc{seL4} currently still trusts address translations; it models page tables as a data structure in regular memory, thus not capturing the possibility that even
% temporarily destroying the mappings and restoring them can actually crash the OS. \textsc{CertiKOS} papers share little in the way of precise details about
% their virtual memory management, but because their core technology is based on a fork of \textsc{CompCert}, whose model of memory is
% a set of unordered block allocations, we can infer their proofs must also trust these translations.
\subsection{Mapping a New Page}
One of the key tasks of a page fault handler in a general-purpose OS kernel is
to map new pages into an address space by writing into an existing page table via a call
\[\textsf{vaspace\_mappage(pte\_t *pml4, void *va,uintptr\_t pa, uint64\_t perm )}\] in Figure \ref{fig:mapping_code}).
To do so, with a given allocated a fresh page (\textsf{pa}), then calculate the appropriate
known-valid page table walks (\textsf{pte\_walkpgdir} Line X in Figure \ref{fig:mapping_code})  and update the appropriate L1 page table entry (permissions and frame address Lines in Figure \ref{fig:mapping_code}).
%\lstset{
%  columns=fullflexible,
%  numbers=left,
%  basicstyle=\ttfamily,
%  keywordstyle=\color{blue}\bfseries,
%  morekeywords={mov,add,call},
%  emph={rsp,rdx,rax,rbx,rbp,rsi,rdi,rcx,r8,r9,r10,r11,r12,r13,r14,r15},
%  emphstyle=\color{green},
%  emph={[2]cr3},
%  emphstyle={[2]\color{violet}},
%  morecomment=[l]{;;},
%  mathescape
%}
\newcommand{\sumwalkabsent}{
  \ownGhost\gammaPred{\authfrag{\singletonMap{\texttt{entry+KERNBASE}}{(\textsf{qfrac}, \textsf{entry})}}}
}

\newcommand{\fpaddr}{\texttt{fpaddr}}
\newcommand{\specline}[1]{{\color{blue}\left\{#1\right\}}}
\newcommand{\sumapacesfull}[2]{
  \ownGhost\gammaPreds{\authfull{\singletonMap{#1}{#2}}}
\todo{ismail change the following spec base on the changes you have in the code}
;;complstatus_t vaspace_mappage(pte_t *pml4, void *va,uintptr_t pa, uint64_t perm ) {
;;setting up the stack      
;;char *a; pte_t *pte;
;;a = (char *)PGROUNDUP((uintptr_t)va);
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,m)  \ast \texttt{rdi}\mapsto_{r} \textsf{plm4} \ast \texttt{rsi}\mapsto_{r} \textsf{va} \ast  \texttt{rdx}\mapsto_{r} \textsf{pa} \ast \texttt{rcx}\mapsto_{r} \textsf{perm} \ast \texttt{rax}\mapsto_{\textsf{r}} \_ \ast \ulcorner \texttt{aligned va} \land  \theta \; !!\; \vaddr = \texttt{None}\urcorner\ast }_{\rtv}$
$\specline{\texttt{rbp-8}\mapsto_{v} \textsf{plm4} \ast \texttt{rbp-16}\mapsto_{v} \textsf{va}  \ast \texttt{rbp-24}\mapsto_{v} \textsf{pa} \ast \texttt{rbp-32}\mapsto_{v} \textsf{perm} \ast \textsf{r8} \mapsto_{\textsf{r}} \textsf{\_}}_{\rtv}$
;;char *a; pte_t *pte;
;;a = (char *)PGROUNDUP((uintptr_t)va);
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,m) \ast  \texttt{rdi}\mapsto_{\textsf{r}} \vaddr  \;\ast}_{\rtv}$
;;pte = walkpgdir_succ(pml4, a);
mov    -0x28[rbp],rsi
mov    0x0,rax
callq  291 <vaspace_mappage+0x41>	28d: R_X86_64_PLT32	walkpgdir_succ-0x4
mov    0x0,r8
mov    rax,rcx
mov    rcx,-0x30[rbp]
;;pte->writable     = PTE_WRITEABLE_SET(perm);
pte->user_acc     = PTE_USER_ACC_SET(perm);
;;pte->writethrough = PTE_WRITETHRU_SET(perm);
;;pte->nocache      = PTE_NOCACHE_SET(perm);
pte->pfn          = PTE_ADDR_TO_PFN(pa);
mov    -0x18[rbp],rcx
shr    0xc,rcx
mov    -0x30[rbp],rdx
mov    [rdx],rsi
movabs 0xffffffffff,rdi
and    rdi,rcx
shl    0xc,rcx
movabs 0xfff0000000000fff,rdi
and    rdi,rsi
or     rcx,rsi
mov    rsi,[rdx]
;;pte->nx           = PTE_NX_SET(perm);
;;pte->present      = 1;
;;clean up the stack and return
  \caption{Specification and proof of code for mapping a new page.}
\label{fig:mapping_code}

In Figure \ref{fig:mapping_code}, we see an address ($\vaddr$) currently not
mapped to a page ($\theta \; !!\; \vaddr = \texttt{None}$). Mapping a fresh
phyiscal page to back the desired virtual page first requires ensuring
the existence of a memory location for an appropriate L1 table entry.
The code uses a helper function \lstinline{pte\_walkpgdir} (discussed again in Section \ref{sec:traversing}).
\textsf{pte\_walkpgdir}'s postcondition contains virtual \emph{PTE} pointsto assertions ($\mapsto_{\textsf{vpte}}$) \todo{colin, these are no more plain physical pointstos}  both for ensuring partial page table walk reaching the
L1 entry (l1e) with the justification of higher levels of the page table exist, and for allowing access to the memory of the L1 entry via virtual address.
A PTE points-to is defined just like the normal virtual points-to of Figure \ref{fig:virtualpointstosharing}, except the physical address (here, \textsf{pa}) is explicit in the assertion 
rather than existentially quantified:
 \[
\begin{array}{l}
    \vaddr\mapsto_{\textsf{vpte,q}} \; \paddr \; \vpage : \mathsf{vProp}~\Sigma \stackrel{\triangle}{=} 
    \exists \delta\ldotp
	(\lambda \mathit{cr3val}\ldotp
	\sumapaces{\mathit{cr3val}}\delta) \ast 
  \sumwalkabs\vaddr\qfrac\paddr \ast \paddr \mapsto_{\mathsf{p}} \vpage
\end{array}
\]
This supports rules for accessing memory
at that virtual address, but exposing the physical location being modified
makes this convenient to use for page table modifications, since we must ensure
the modified data is at the specific physical location that will be used as the L1 entry.
This is also guaranteed ($ \ulcorner
\texttt{addr\_L1}(\vaddr,\entryo) = \paddr\urcorner$).
% After obtaining a virtual address \textsf{pte\_addr} in \textsf{rax} backed 
% by the physical memory for the L1 entry that will be used to translate the virtual addresses
% we are mapping, we save it to \textsf{r14} to be updated later in Line 9.

The crucial step is actually updating the L1 entry (Line X\todo{X} in Figure \ref{fig:mapping_code}),
via the virtual address
(\textsf{pte\_addr}) known to translate to the appropriate physical address, in our example the L1
table entry address ($\textsf{addr\_L1}(\textsf{va, l1e})$), to hold the freshly
allocated physical page address (\textsf{fpaddr}) in Line X.
After this write, we have (Specification Line Y)
\[\texttt{pte\_addr} \mapsto_{\texttt{vpte}}  \; \paddr\; \textsf{fpaddr}  \qquad\qquad (\textsf{by the PTE variant of Rule}~\TirNameStyle{WriteToVirtMemFromReg})\]
